---
title: "ETC3250/5250 Tutorial 7 Instructions"
subtitle: "Forests and hyperplanes"
author: "prepared by Professor Di Cook"
date: "Week 7"
output:
  html_document:
    after_body: tutorial-footer.html
    css: tutorial.css
---

```{r, echo = FALSE, message = FALSE, warning = FALSE, warning = FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  error = FALSE, 
  echo = FALSE,
  eval = FALSE,
  collapse = TRUE,
  comment = "#",
  fig.height = 4,
  fig.width = 8,
  fig.align = "center",
  cache = FALSE
)
```


# 🎯 Objective

The objectives for this week are to 

* Learn about fitting a forest model and use the diagnostics 
* Compare the classification tree with a forest
* Understand the the idea of bagging to refine model fitting
* Learn about fitting a support vector machine model 
* Examine the differences between results from different models


# 💻 Group discussion 

Estimating the maximal margin hyperplane corresponds to maximising (believe me, or work your way through [Sidharth's page](https://medium.com/analytics-vidhya/math-behind-support-vector-machines-642421e45b08))

$$\sum_i{\alpha_i} - \frac12 \sum_i\sum_k\alpha_i\alpha_ky_iy_k\mathbf{x}_i^T\mathbf{x}_k$$
where $\mathbf{x}_i, \mathbf{x}_k$ are two $p$-dimensional data vectors (observations), and the coefficients of the separating hyperplane

$$\beta_0 + \beta_1 x_1 + \dots + \beta_p x_p = 0$$

are computed from the support vectors and weights $\alpha$'s from the optimisation as follows:

$$\mathbf{\beta}_j=\sum_{i=1}^s (\alpha_iy_i)\mathbf{x}_{ij}$$

Now, $i, k = 1, ..., n$ but because only some observations are used to compute $\beta$ most are 0, and we can sum only over $1, ..., s$, where $s$ is the number of support vectors. 

With the kernel trick, 

$$\sum_i{\alpha_i} - \frac12\sum_i\sum_k\alpha_i\alpha_ky_iy_kK(\mathbf{x}_i^T\mathbf{x}_k)$$

Try one kernel function transformation to show that $K(\mathbf{x}_i^T\mathbf{x}_k) = \psi(\mathbf{x}_i)^T\psi(\mathbf{x}_k)$. You can think of $\psi()$ as transformations of the predictors, $\mathbf{x}$.

Fill in the steps to go from the first line to the last. Note that $p=2$. (You can find all the steps in the lecture notes.)

$$
\begin{align*}

\mathcal{K}(\mathbf{x_i}, \mathbf{x_k}) 
                                    & = (1 + \langle \mathbf{x_i}, \mathbf{x_k}\rangle) ^2 \\
                                    & = \left(1 + \sum_{j = 1}^2 x_{ij}x_{kj} \right) ^2 \\
                                    & = (1 + x_{i1}x_{k1} + x_{i2}x_{k2})^2 \\
                                    & = (1, x_{i1}^2, x_{i2}^2, \sqrt2x_{i1}, \sqrt2x_{i2}, \sqrt2x_{i1}x_{i2})^T(1, x_{k1}^2, x_{k2}^2, \sqrt2x_{k1},                                         \sqrt2x_{k2}, \sqrt2x_{k1}x_{k2}) \\
                                    & = \langle \psi(\mathbf{x_i}), \psi(\mathbf{x_k}) \rangle
\end{align*}
$$
Have a chat about why this algebraic "trick" is neat.

$\langle \mathbf{x_i}, \mathbf{x_k}\rangle = \sum^{p}_{j=1}x_{j}y_{j}$

* This is doing a non-linear model fitting, but it's all inside the mathematics
* Kernel trick is an elegant mathematical term to fit non-linear terms to the data.

# 👋 Getting started

If you are in a zoom tutorial, say hello in the chat. If in person, do say hello to your tutor and to your neighbours. 


# Exercise 1

Effect of different subsets of variables, and feature engineering

## a. 

Fit the *tree* to olive oils, using a training split of 2/3, using only regions 2, 3,  and  the predictors linoleic and arachidic. Report the balanced accuracy of the test set, and make a plot of the boundary.

```{r setup, include=FALSE}
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(patchwork)

options(digits = 4)

# Set default theme for document
theme_set(theme_bw())
```

```{r}
olive <- read_csv("http://www.ggobi.org/book/data/olive.csv") %>%
  rename(name = `...1`) 

notsouth <- olive %>%
  # Regions 2 and 3 only
  filter(region != 1) %>%
  select(region, linoleic, arachidic) %>%
  # Convert response to categorical variables
  mutate(region = factor(region))
```

```{r}
set.seed(2021)

# Split data into training & test set
notsouth_split <- initial_split(notsouth, prop = 2/3, strata = region)
notsouth_tr <- training(notsouth_split)
notsouth_ts <- testing(notsouth_split)

# Fit classification tree on training set
notsouth_tree <- decision_tree() %>%
  set_engine("rpart") %>% 
  set_mode("classification") %>%
  fit(region ~ ., notsouth_tr)
```

```{r}
notsouth_tree %>% 
  extract_fit_engine() %>% 
  prp(type = 3, ni = TRUE, nn = TRUE, extra = 2, box.palette = "RdBu")

ggplot(data = notsouth_tr, 
       aes(x = linoleic,
           y = arachidic,
           colour = region)) +
  geom_point() + 
  scale_color_brewer("", palette="Dark2") + 
  geom_vline(xintercept = 1054) +
  theme(aspect.ratio = 1)
```


```{r}
# Create dataset that is close to original dataset
notsouth_p <- 
  as_tibble(expand.grid(linoleic = seq(440, 1500, 10), 
                        arachidic = seq(0, 105, 2))) 

# Predict using classification tree built
notsouth_p  <- notsouth_p  %>%
  mutate(region_tree = predict(notsouth_tree, notsouth_p )$.pred_class)

ggplot() +
  # Predictions
  geom_point(data = notsouth_p, 
             aes(x = linoleic, 
                 y = arachidic, 
                 color = region_tree), 
             alpha = 0.1) +
  # Overlay with actual dataset
  geom_point(data = notsouth, 
             aes(x = linoleic, 
                 y = arachidic, 
                 color = region, 
                 shape = region)) +
  scale_color_brewer("", palette = "Dark2") +
  theme_bw() + theme(aspect.ratio = 1, legend.position = "none") +
  ggtitle("tree")
```

```{r}
# Predictions for test set
notsouth_ts_pred <- notsouth_ts %>%
  mutate(pred_tree = predict(notsouth_tree, notsouth_ts)$.pred_class)

# Balanced accuracy
yardstick::bal_accuracy(data = notsouth_ts_pred, truth = region, estimate = pred_tree)

# Confusion matrix
yardstick::conf_mat(notsouth_ts_pred, truth = region, estimate = pred_tree)
```

**The balanced accuracy is** `r bal_accuracy(notsouth_ts_pred, region, pred_rf)$.estimate`.

## b.

Fit a random forest to the full data, using only `linoleic` and `arachidic` as predictors, report the balanced accuracy for the test set, and make a plot of the boundary.

```{r}
# Fit classification tree on training set
notsouth_rf <- 
  rand_forest() %>% 
  set_engine("randomForest", 
             importance = TRUE,
             proximity = TRUE) %>% 
  set_mode("classification") %>% 
  fit(region ~ ., data = notsouth_tr) 

notsouth_rf
```

* See `?randomForest()`: 
  * `mtry` = No. of variables used at each decision node - $\sqrt{\text{total no. of variables}}$
  * `trees` = Number of trees in random forest - $500$ by default
* OOB error estimate tells us that 100% of OOB samples are correctly classified
  
```{r}
# Predict using hypothetical dataset created in (a) 
notsouth_p  <- notsouth_p  %>%
  mutate(region_rf = predict(notsouth_rf, notsouth_p)$.pred_class)

ggplot() +
  # Predictions
  geom_point(data = notsouth_p, 
             aes(x = linoleic, 
                 y = arachidic, 
                 color = region_rf), 
             alpha = 0.1) +
  # Overlay with actual dataset
  geom_point(data = notsouth, 
             aes(x = linoleic,
                 y = arachidic, 
                 color = region,
                 shape = region)) +
  scale_color_brewer("", palette = "Dark2") +
  theme_bw() + 
  theme(aspect.ratio = 1, legend.position = "none") +
  ggtitle("forest")
```

```{r}
# Predictions for test set
notsouth_ts_pred <- notsouth_ts %>%
  mutate(pred_rf = predict(notsouth_rf, notsouth_ts)$.pred_class)

# Balanced accuracy
yardstick::bal_accuracy(data = notsouth_ts_pred, truth = region, estimate = pred_rf)

# Confusion matrix
yardstick::conf_mat(notsouth_ts_pred, truth = region, estimate = pred_rf)
```

## c. 

Explain the difference between the single tree and random forest boundaries.

* The forest model is ever so slightly curved around the two clusters, whereas the tree is a single split. 
* *Note*: with a different seed, the boundary is sometime boxy at the higher values of `arachidic`. 
  * Using a larger number of trees in the forest might stabilise the result.

## d. 

Fit the random forest again to the full set of variables, and  compute the variable importance. Describe the order of importance of variables.

```{r}
notsouth_all <- olive %>%
  # Only regions 2 & 3
  filter(region != 1) %>%
  # Full set of variables
  select(region, palmitic:arachidic) %>%
  mutate(region = factor(region))
```

```{r}
# Split data into training & test set
set.seed(2021)
notsouth_all_split <- initial_split(notsouth_all, prop = 2/3, strata = region) 
notsouth_all_tr <- training(notsouth_all_split)
notsouth_all_ts <- testing(notsouth_all_split)

# Fit random forest to training set
notsouth_all_rf <-
  rand_forest() %>% 
  set_engine("randomForest", 
             importance = TRUE,
             proximity = TRUE) %>% 
  set_mode("classification") %>% 
  fit(region ~ ., data = notsouth_all_tr)

notsouth_all_rf
```

```{r}
notsouth_all_rf$fit$importance
```

* The most important variables by far are linoleic and oleic, with arachidic much less important.

## e. 

Create a new variable called `linoarch` that is $0.377 \times linoleic + 0.926\times arachidic$. Make a plot of this variable against `arachidic`. Fit the tree model to the same training data using this variable in addition to `linoleic` and `arachidic`. Why doesn't the tree use this new variable? It has a bigger difference between the two groups than `linoleic`? Change the order of the variables, so that `linoarch` is before `linoleic` and re-fit the tree. Does it use this variable now? Why do you think this is?

```{r}
# Create new variable onto training set in (a)
notsouth_tr <- notsouth_tr %>%
  mutate(linoarch = 0.377 * linoleic + 0.926 * arachidic)

# Arachidic vs. linoleic
p1 <- 
  ggplot() +
  geom_point(data = notsouth_tr, 
             aes(x = linoleic, 
                 y = arachidic, 
                 color = region, 
                 shape = region)) +
  scale_color_brewer("", palette="Dark2") +
  theme_bw() + 
  theme(aspect.ratio = 1,
        legend.position = "none") +
  ggtitle("tree")

# Arachidic vs linoarch
p2 <- ggplot() +
  geom_point(data = notsouth_tr, 
             aes(x = linoarch, 
                 y = arachidic, 
                 color = region, 
                 shape = region)) +
  scale_color_brewer("", palette="Dark2") +
  theme_bw() + 
  theme(aspect.ratio = 1,
        legend.position = "none") +
  ggtitle("tree")

p1 + p2
```

```{r}
# Fit classification tree
notsouth_tree <- decision_tree() %>%
  set_engine("rpart") %>% 
  set_mode("classification") %>%
  fit(region ~ ., notsouth_tr)
 
notsouth_tree %>% 
  extract_fit_engine() %>% 
  prp(type = 3, ni = TRUE, nn = TRUE, extra = 2, box.palette = "RdBu")
```

```{r}
# Change the order of the variables - linoarch is before linoleic
notsouth_tree <- decision_tree() %>%
  set_engine("rpart") %>% 
  set_mode("classification") %>%
  fit(region ~ linoarch + linoleic + arachidic, notsouth_tr)

notsouth_tree %>% 
  extract_fit_engine() %>% 
  prp(type = 3, ni = TRUE, nn = TRUE, extra = 2, box.palette = "RdBu")
```

* The new variable is seen when order is changed. 
* The initial fit doesn't see the better variable because both variables have a split with the *same impurity value*, and thus the first variable entered into the model is the one that is selected. 
* We can see that the new variable is "better" than the first because there is a bigger gap between the two groups, but this isn't a factor considered by a tree model.

## f. 

Fit the random forest again to the full set of variables, including linoarch and  compute the variable importance. Describe the order of importance of variables. Does the forest see the new variable?

```{r}
notsouth_rf <- rand_forest() %>%
  set_engine("randomForest",
             importance = TRUE, proximity = TRUE) %>%
  set_mode("classification") %>%
  fit(region ~ ., data = notsouth_tr) 

notsouth_rf$fit$importance
```

* Yes, the forest sees the new variable. 
* However, it considers it to be equally as important as linoleic. 
* The forest also doesn't recognise that the new variable is better because it also is not considering the magnitude of the gap between groups.

# Exercise 2

Support vector machine model fitting

## a. 

Fit the linear SVM to olive oils, using a training split of 2/3, using only regions 2, 3, and the predictors linoleic and arachidic. Report the training and test error, list the support vectors, the coefficients for the support vectors and the equation for the separating hyperplane, and
$$???\times\text{linoleic}+???\times\text{arachidic}+??? > 0$$ and make a plot of the boundary.

```{r}
notsouth <- olive %>%
  filter(region != 1) %>%
  select(region, linoleic, arachidic) %>%
  mutate(region = factor(region)) %>%
  mutate(across(where(is.numeric), ~ (.x - mean(.x)) / sd(.x)))
```

```{r}
set.seed(2021)
notsouth_split <- initial_split(notsouth, prop = 2/3, strata = region)
notsouth_tr <- training(notsouth_split)
notsouth_ts <- testing(notsouth_split)

library(kernlab)

svm_mod <-
  svm_rbf(cost = 10) %>%
  set_mode("classification") %>%
  set_engine("kernlab", 
             kernel = "vanilladot", # linear kernel, see ?kernlab::ksvm()
             scaled = FALSE)

notsouth_svm <- svm_mod %>%
  fit(region ~ ., data = notsouth_tr)
```

```{r}
notsouth_p <- as_tibble(expand_grid(linoleic = seq(-2.2, 2.2, 0.1), 
                                    arachidic = seq(-2, 2, 0.1))) 

notsouth_p  <- notsouth_p  %>%
  mutate(region_svm = predict(notsouth_svm, notsouth_p )$.pred_class)

ggplot() +
  # Predicted values
  geom_point(data = notsouth_p, 
             aes(x = linoleic, 
                 y = arachidic,
                 color = region_svm), 
             alpha = 0.1) +
  # Overlay with actual data
  geom_point(data = notsouth, 
             aes(x = linoleic, 
                 y = arachidic, 
                 color = region, 
                 shape = region)) +
  # Circle the support vectors
  geom_point(data = notsouth_tr %>% slice(notsouth_svm$fit@SVindex), # Extract support vectors
             aes(x = linoleic, y = arachidic), 
             shape = 1, 
             size = 3,
             colour = "black") +
  scale_color_brewer("", palette="Dark2") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  ggtitle("SVM") +
  geom_abline(intercept = 1.45396, slope = -3.478113)
```

The $\alpha$'s, indexes of support vectors and $\beta_0$, and the observations that are the support vectors are:

```{r}
notsouth_svm$fit@coef # \alpha * y_i
notsouth_svm$fit@SVindex # Indexes (row numbers) of support vectors
notsouth_svm$fit@b # Negative intercept term -b_0

# Extract support vectors using indexes
notsouth_tr[notsouth_svm$fit@SVindex, ] # Support vectors; notsouth_tr %>% slice(6,102,132)
notsouth_tr$region[notsouth_svm$fit@SVindex] # Response variable of support vectors
```

Use the formula (see *lecture-07b slide 22*)

$$\mathbf{\beta}_j=\sum_{i=1}^s (\alpha_iy_i)\mathbf{x}_{ij}$$

to compute the remaining coefficients.


$$\beta_1 = -6.1337 * 0.614 + 3.2251 * (-0.348) + 2.9085 * 0.5234 = -3.366118$$


$$\beta_2 = -6.1337 * 0.351 + 3.225187 * 1.63 + 2.9085 * (-1.40) = -0.9678$$

which would give the equation of the separating hyperplane to be

$$  1.4072 + -3.366118\mbox{linoleic}  - 0.9678\mbox{arachidic} = 0$$

and rearranging for arachidic as the focus gives

$$\mbox{arachidic} = 1.4072/0.9678  - 3.366118/0.9678\mbox{linoleic}$$

slope of the line is -3.478113, and intercept is 1.45396, gives the line drawn on the plot above.


## b. 

Fit a radial kernel SVM, with a variety of cost values, to examine the effect on the boundary. (Try changing the cost from 0.3 up to 10.)

```{r}
svm_mod2 <-
  svm_rbf(cost = 30) %>%
  set_mode("classification") %>%
  set_engine("kernlab",
             kernel = "rbfdot", # Radial basis kernel
             scaled = FALSE)

notsouth_svm2 <- svm_mod2 %>%
  fit(region ~ ., data = notsouth_tr)

notsouth_p  <- notsouth_p  %>%
  mutate(region_svm2 = predict(notsouth_svm2, notsouth_p)$.pred_class)

ggplot() +
  geom_point(data = notsouth_p,
             aes(x = linoleic, 
                 y = arachidic,
                 color = region_svm2),
             alpha = 0.3) +
  geom_point(data = notsouth,
             aes(x = linoleic,
                 y = arachidic,
                 color = region,
                 shape = region)) +
  scale_color_brewer("", palette = "Dark2") +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  ggtitle("SVM")
```

* The boundary is always a small circle wrapping the smaller group. 
* Very low values of cost break the fit, though, and give poor prediction of the smaller group.

* Variance/bias trade-off
  * Higher cost - Higher variance
  * Lower cost - Higher bias

##### © Copyright 2021 Monash University


